---
title: "Customer Behavior and Product Performance Analysis in Fashion E-Commerce: A Study Using H&M Transaction Data"
author: "Tanisha Magikar, Nikhil Kumar, Bhavika Rathi, Favour Forchu"
date: "Today"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

---

# Research Topic

**Customer Behavior and Product Performance Analysis in Fashion E-Commerce: A Study Using H&M Transaction Data**

---

# Team Members

| Name             | Role                        |
|------------------|-----------------------------|
| Tanisha Magikar  | Team Member                 |
| Nikhil Kumar     | Team Member                 |
| Bhavika Rathi    | Team Member                 |
| Favour Forchu    | Team Member                 |

---

# Data Description

## Dataset

We are using the **H&M Personalized Fashion Recommendations** dataset, publicly available on Kaggle as part of a competition hosted by H&M Group.

**Dataset Link:** [https://www.kaggle.com/competitions/h-and-m-personalized-fashion-recommendations/data](https://www.kaggle.com/competitions/h-and-m-personalized-fashion-recommendations/data)

The dataset contains three primary files:

- **`transactions_train.csv`** — Purchase-level records including customer ID, article ID, price, and purchase date.
- **`articles.csv`** — Article (product) metadata including product type, group name, department name, color, and more.
- **`customers.csv`** — Customer-level attributes including age, membership status, and postal code.

```{r load-libraries}
# Load required libraries
library(tidyverse)   # Data manipulation and visualization
library(lubridate)   # Date handling
library(scales)      # Axis formatting in plots
library(knitr)       # Table rendering
```

```{r load-data, eval=FALSE}
# NOTE: Data files must be downloaded from Kaggle and placed in a /data subdirectory
# Set eval=TRUE once data is available locally

transactions <- read_csv("data/transactions_train.csv")
articles     <- read_csv("data/articles.csv")
customers    <- read_csv("data/customers.csv")
```

```{r mock-preview, echo=FALSE}
# Illustrative structure — replace with real data once loaded
transactions_mock <- tibble(
  t_dat        = as.Date(c("2020-09-07", "2020-09-08", "2020-09-09")),
  customer_id  = c("C001", "C002", "C001"),
  article_id   = c("A101", "A202", "A303"),
  price        = c(29.99, 49.95, 15.00),
  sales_channel_id = c(2L, 1L, 2L)
)

kable(head(transactions_mock), caption = "Example: transactions_train.csv structure")
```

---

## Anticipated Data Cleaning Steps

The following cleaning steps are planned before analysis begins:

1. **Handle missing values** — Inspect each data frame for `NA`s. Decide on a per-column basis whether to drop rows (e.g., missing `customer_id`) or impute (e.g., median imputation for `age`).

2. **Remove duplicates** — Deduplicate rows in all three files, particularly in `transactions_train.csv` where repeated purchase records could inflate counts.

3. **Convert data types** — Parse `t_dat` to `Date`, ensure `article_id` and `customer_id` are character strings, and convert relevant columns to factors.

4. **Merge datasets and standardize categorical variables** — Join `transactions` with `articles` and `customers` on their shared keys. Standardize categorical columns (e.g., trim whitespace, unify casing) to avoid spurious category splits.

5. **Handle outliers** — After exploratory analysis, examine extreme values in `price`, transaction frequency, and customer age. Apply domain-appropriate treatment (e.g., winsorizing prices, removing clearly erroneous ages).

6. **Feature engineering** — Derive new columns to support the research questions, including:
   - `total_purchases_per_customer` — Transaction count per customer.
   - `recency` — Days since a customer's most recent purchase.
   - `loyalty_tier` — Bucketed customer segments (one-time, occasional, regular, loyal).
   - `sell_through_velocity` — Units sold divided by days the product was active.

```{r cleaning-demo, eval=FALSE}
# Example cleaning pipeline (illustrative — run after data is loaded)

transactions_clean <- transactions %>%
  mutate(t_dat = as.Date(t_dat)) %>%
  distinct() %>%
  filter(!is.na(customer_id), !is.na(article_id))

customers_clean <- customers %>%
  mutate(age = as.numeric(age)) %>%
  filter(age >= 16, age <= 100)   # Remove implausible ages

articles_clean <- articles %>%
  mutate(across(where(is.character), str_trim))
```

---

## Basic / Marginal Summaries Planned

- Distribution of customer ages (histogram)
- Overall transaction volume over time (line chart)
- Top 10 most purchased product departments (bar chart)
- Summary statistics for `price` (mean, median, IQR)
- Count of unique customers vs. unique articles purchased

---

# Project Idea / Research Questions

## What Are We Trying to Investigate?

We aim to understand **how customers interact with fashion products** in an e-commerce setting — specifically looking at purchasing patterns, customer loyalty, and product performance. The four core research questions are:

> **RQ1:** What items are frequently bought together? *(Cross-sell / product affinity)*

> **RQ2:** What does the age distribution of customers look like, and do different age groups buy different product types?

> **RQ3:** What is the repeat purchase rate? How many customers buy once versus becoming regulars?

> **RQ4:** Which items sell quickly versus sitting in inventory? *(Inventory velocity)*

---

## Analyses and Visualizations

### RQ1 — Product Affinity (Co-purchase Analysis)

We will count how often pairs of product categories appear together in the **same customer's overall purchase history**. The results will be displayed as a **co-purchase heatmap**, where both axes represent product departments and fill color encodes co-purchase frequency.

```{r affinity-mockup, fig.cap="Illustrative co-purchase heatmap (placeholder data)", fig.width=7, fig.height=5}
# Placeholder heatmap with synthetic data
set.seed(42)
depts  <- c("Tops", "Bottoms", "Dresses", "Outerwear", "Accessories", "Footwear")
mat    <- matrix(sample(0:500, length(depts)^2, replace = TRUE), nrow = length(depts),
                 dimnames = list(depts, depts))
diag(mat) <- 0

as.data.frame(as.table(mat)) %>%
  rename(Dept1 = Var1, Dept2 = Var2, CoCount = Freq) %>%
  ggplot(aes(x = Dept1, y = Dept2, fill = CoCount)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#E8002D",
                      name = "Co-purchases") +
  labs(title    = "Product Department Co-purchase Heatmap (Illustrative)",
       subtitle = "Higher values = more frequently bought together",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

### RQ2 — Age Distribution and Product Preferences

Customers will be binned into age groups (e.g., 16–24, 25–34, 35–49, 50–64, 65+). We will use:

- A **histogram** of the overall age distribution.
- A **heatmap** showing which product departments each age group over-indexes in (normalized purchase share).

```{r age-mockup, fig.cap="Illustrative age distribution (placeholder data)", fig.width=6, fig.height=4}
set.seed(7)
age_data <- tibble(age = round(rnorm(5000, mean = 38, sd = 14))) %>%
  filter(age >= 16, age <= 85)

ggplot(age_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#E8002D", color = "white", alpha = 0.85) +
  scale_x_continuous(breaks = seq(15, 85, 10)) +
  labs(title    = "Customer Age Distribution (Illustrative)",
       subtitle = "Placeholder — will be replaced with real H&M data",
       x = "Age", y = "Number of Customers") +
  theme_minimal(base_size = 12)
```

---

### RQ3 — Repeat Purchase Rate / Loyalty Tiers

Customers will be classified into **loyalty tiers** based on total transaction counts:

| Tier       | Definition                   |
|------------|------------------------------|
| One-time   | Exactly 1 purchase           |
| Occasional | 2–5 purchases                |
| Regular    | 6–15 purchases               |
| Loyal      | 16+ purchases                |

Side-by-side bar charts will compare the **share of customers** vs. **share of revenue** per tier — to show whether loyal customers are disproportionately valuable.

```{r loyalty-mockup, fig.cap="Illustrative loyalty tier breakdown (placeholder)", fig.width=7, fig.height=4}
loyalty_mock <- tibble(
  Tier            = factor(c("One-time", "Occasional", "Regular", "Loyal"),
                           levels = c("One-time", "Occasional", "Regular", "Loyal")),
  Pct_Customers   = c(48, 30, 15, 7),
  Pct_Revenue     = c(18, 28, 30, 24)
) %>%
  pivot_longer(cols = c(Pct_Customers, Pct_Revenue),
               names_to  = "Metric",
               values_to = "Percentage") %>%
  mutate(Metric = recode(Metric,
                         "Pct_Customers" = "% of Customers",
                         "Pct_Revenue"   = "% of Revenue"))

ggplot(loyalty_mock, aes(x = Tier, y = Percentage, fill = Metric)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.88) +
  scale_fill_manual(values = c("% of Customers" = "#E8002D",
                                "% of Revenue"   = "#222222")) +
  labs(title    = "Customer Loyalty Tiers: Share of Customers vs. Revenue (Illustrative)",
       subtitle = "Placeholder — will be replaced with real H&M data",
       x = NULL, y = "Percentage (%)", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

---

### RQ4 — Inventory Velocity (Fast vs. Slow-moving Products)

For each product, we will compute a **sell-through velocity** = units sold / days active in the dataset. A **scatter plot** of units sold (y-axis) vs. days active (x-axis), colored by product category, will reveal fast-moving vs. stagnant inventory.

```{r velocity-mockup, fig.cap="Illustrative sell-through velocity scatter plot (placeholder)", fig.width=7, fig.height=5}
set.seed(99)
n     <- 200
cats  <- c("Tops", "Bottoms", "Dresses", "Outerwear", "Accessories")
vel_mock <- tibble(
  days_active   = sample(1:365, n, replace = TRUE),
  units_sold    = pmax(1, round(runif(n, 1, 500) - days_active * 0.3 + rnorm(n, 50, 30))),
  category      = sample(cats, n, replace = TRUE)
)

ggplot(vel_mock, aes(x = days_active, y = units_sold, color = category)) +
  geom_point(alpha = 0.65, size = 2.2) +
  scale_color_brewer(palette = "Set1") +
  labs(title    = "Product Sell-Through Velocity (Illustrative)",
       subtitle = "Placeholder — will be replaced with real H&M data",
       x = "Days Active in Dataset",
       y = "Total Units Sold",
       color = "Category") +
  theme_minimal(base_size = 12)
```

---

# Summary

This project investigates customer behavior and product performance in fashion e-commerce using H&M's transaction data. Our four research questions span cross-selling strategy, demographic segmentation, customer loyalty, and inventory management. All analysis will be conducted in **R** using `tidyverse` and `ggplot2`, and delivered as a fully reproducible R Markdown report.

---

*Report prepared for STAT / DS Course — Spring 2026*
