---
title: "Customer Behavior and Product Performance Analysis in Fashion E-Commerce: A Study Using H&M Transaction Data"
author: "Tanisha Magikar, Nikhil Kumar, Bhavika Rathi, Favour Forchu"
date: "May 04, 2026"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## Research Topic

**Customer Behavior and Product Performance Analysis in Fashion E-Commerce: A Study Using H&M Transaction Data**

***

## Team Members

- Tanisha Magikar
- Nikhil Kumar
- Bhavika Rathi
- Favour Forchu

## Data Description

### Dataset

We are using the **H&M Personalized Fashion Recommendations** dataset, publicly available on Kaggle as part of a competition hosted by H&M Group.

**Dataset Link:** [https://www.kaggle.com/competitions/h-and-m-personalized-fashion-recommendations/data](https://www.kaggle.com/competitions/h-and-m-personalized-fashion-recommendations/data)

The dataset contains three primary files:

- **`transactions_train.csv`** — Purchase-level records including customer ID, article ID, price, and purchase date.
- **`articles.csv`** — Article (product) metadata including product type, group name, department name, color, and more.
- **`customers.csv`** — Customer-level attributes including age, membership status, and postal code.

```{r load-libraries}
library(tidyverse)   
library(lubridate)  
library(scales)      
library(knitr)       
```

```{r load-data, eval=TRUE}

transactions <- read_csv("data/transactions_train.csv", show_col_types = FALSE)
articles     <- read_csv("data/articles.csv", show_col_types = FALSE)
customers    <- read_csv("data/customers.csv", show_col_types = FALSE)
```


***

### Anticipated Data Cleaning Steps

The following cleaning steps are planned before analysis begins:

1. **Handle missing values** — Inspect each data frame for `NA`s. Decide on a per-column basis whether to drop rows (e.g., missing `customer_id`) or impute (e.g., median imputation for `age`).

2. **Remove duplicates** — Deduplicate rows in all three files, particularly in `transactions_train.csv` where repeated purchase records could inflate counts.

3. **Convert data types** — Parse `t_dat` to `Date`, ensure `article_id` and `customer_id` are character strings, and convert relevant columns to factors.

4. **Merge datasets and standardize categorical variables** — Join `transactions` with `articles` and `customers` on their shared keys. Standardize categorical columns (e.g., trim whitespace, unify casing) to avoid spurious category splits.

5. **Handle outliers** — After exploratory analysis, examine extreme values in `price`, transaction frequency, and customer age. Apply domain-appropriate treatment (e.g., winsorizing prices, removing clearly erroneous ages).

6. **Feature engineering** — Derive new columns to support the research questions, including:
   - `total_purchases_per_customer` — Transaction count per customer.
   - `recency` — Days since a customer's most recent purchase.
   - `loyalty_tier` — Bucketed customer segments (one-time, occasional, regular, loyal).
   - `sell_through_velocity` — Units sold divided by days the product was active.

```{r cleaning-demo, eval=TRUE}


colSums(is.na(transactions))
colSums(is.na(customers))
colSums(is.na(articles))

# Clean transactions
transactions_clean <- transactions %>%
  mutate(
    t_dat = as.Date(t_dat),
    customer_id = as.character(customer_id),
    article_id = as.character(article_id)
  ) %>%
  distinct() %>%
  filter(
    !is.na(customer_id),
    !is.na(article_id),
    !is.na(t_dat),
    !is.na(price)
  )

# Clean customers
customers_clean <- customers %>%
  mutate(
    customer_id = as.character(customer_id),
    age = as.numeric(age)
  ) %>%
  distinct() %>%
  filter(is.na(age) | between(age, 16, 100))

# Clean articles
articles_clean <- articles %>%
  mutate(
    article_id = as.character(article_id),
    across(where(is.character), str_trim)
  ) %>%
  distinct()

# Customer-level feature engineering
reference_date <- max(transactions_clean$t_dat, na.rm = TRUE)

customer_features <- transactions_clean %>%
  group_by(customer_id) %>%
  summarise(
    total_purchases_per_customer = n(),
    recency = as.numeric(reference_date - max(t_dat, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    loyalty_tier = case_when(
      total_purchases_per_customer == 1 ~ "one-time",
      total_purchases_per_customer <= 5 ~ "occasional",
      total_purchases_per_customer <= 20 ~ "regular",
      TRUE ~ "loyal"
    )
  )

# Product-level feature engineering
article_features <- transactions_clean %>%
  group_by(article_id) %>%
  summarise(
    total_units_sold = n(),
    first_sale_date = min(t_dat, na.rm = TRUE),
    last_sale_date = max(t_dat, na.rm = TRUE),
    active_days = as.numeric(last_sale_date - first_sale_date) + 1,
    .groups = "drop"
  ) %>%
  mutate(
    sell_through_velocity = total_units_sold / active_days
  )

# Avoid building one giant joined table in memory for the full transaction history
glimpse(customer_features)
glimpse(article_features)
```

***

## Basic / Marginal Summaries Planned

- Distribution of customer ages (histogram)
- Overall transaction volume over time (line chart)
- Top 10 most purchased product departments (bar chart)
- Summary statistics for `price` (mean, median, IQR)
- Count of unique customers vs. unique articles purchased

***

## Project Idea / Research Questions

### What Are We Trying to Investigate?

We aim to understand **how customers interact with fashion products** in an e-commerce setting — specifically looking at purchasing patterns, customer loyalty, and product performance. The four core research questions are:

> **RQ1:** What items are frequently bought together? 

> **RQ2:** What does the age distribution of customers look like, and do different age groups buy different product types?

> **RQ3:** What is the repeat purchase rate? How many customers buy once versus becoming regulars?

> **RQ4:** Which items sell quickly versus sitting in inventory? 

***
## Main Skepticism and limitations

This analysis is limited by the structure of the dataset. The data we get is only the completed transactions and we have no ability to capture the browsing, abandoned carts, or untapped demand. Also, the lack of missing or incomplete demographic data (e.g., age) can also bias the outcomes of customer segmentation.

The sell-through velocity measure is also an approximation, as actual inventory status and inventory availability is not given. Consequently, certain products might not seem to move swiftly merely because of low supply and not low demand.

## Analyses and Visualizations

### RQ1 — Product Affinity (Co-purchase Analysis)

We will count how often pairs of product categories appear together in the **same customer's overall purchase history**. The results will be displayed as a **co-purchase heatmap**, where both axes represent product departments and fill color encodes co-purchase frequency.

```{r affinity-real, fig.width=7, fig.height=5}
tx_dept <- transactions_clean %>%
  left_join(articles_clean %>% select(article_id, department_name),
            by = "article_id")
top_depts <- tx_dept %>%
  count(department_name, sort = TRUE) %>%
  slice_head(n = 8) %>%
  pull(department_name)

copurchase <- tx_dept %>%
  filter(department_name %in% top_depts) %>%
  select(customer_id, department_name) %>%
  distinct() %>%
  mutate(value = 1L)

co_matrix <- xtabs(value ~ customer_id + department_name, data = copurchase)
co_counts <- as.matrix(t(co_matrix) %*% co_matrix)
diag(co_counts) <- 0

copurchase <- as_tibble(as.data.frame(as.table(co_counts))) %>%
  rename(
    department_name.x = department_name,
    department_name.y = department_name.1,
    CoCount = Freq
  ) %>%
  filter(CoCount > 0)

ggplot(copurchase, aes(x = department_name.x, y = department_name.y, fill = CoCount)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#E8002D", name = "Co-purchases") +
  labs(title = "Product Department Co-purchase Heatmap",
       subtitle = "How often departments appear together in a customer's history",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

***

### RQ2 — Age Distribution

Customers will be binned into age groups (e.g., 16–24, 25–34, 35–49, 50–64, 65+). We will use:

- A **histogram** of the overall age distribution to see which age group buys the most products.



```{r age-real, fig.width=6, fig.height=4}
ggplot(customers_clean, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#E8002D", color = "white", alpha = 0.85) +
  scale_x_continuous(breaks = seq(15, 85, 10)) +
  labs(title = "Customer Age Distribution",
       x = "Age", y = "Number of Customers") +
  theme_minimal(base_size = 12)
```


***

### RQ3 — Repeat Purchase Rate / Loyalty Tiers

Customers will be classified into **loyalty tiers** based on total transaction counts:

| Tier       | Definition                   |
|------------|------------------------------|
| One-time   | Exactly 1 purchase           |
| Occasional | 2–5 purchases                |
| Regular    | 6–15 purchases               |
| Loyal      | 16+ purchases                |

Side-by-side bar charts will compare the **share of customers** vs. **share of revenue** per tier — to show whether loyal customers are disproportionately valuable.

```{r loyalty-real, fig.width=7, fig.height=4}
loyalty_real <- transactions_clean %>%
  count(customer_id, name = "n_purchases") %>%
  mutate(Tier = case_when(
    n_purchases == 1  ~ "One-time",
    n_purchases <= 5  ~ "Occasional",
    n_purchases <= 15 ~ "Regular",
    TRUE              ~ "Loyal"
  ))

tier_revenue <- transactions_clean %>%
  left_join(loyalty_real, by = "customer_id") %>%
  group_by(Tier) %>%
  summarise(total_revenue = sum(price, na.rm = TRUE))

tier_summary <- loyalty_real %>%
  count(Tier, name = "n_customers") %>%
  left_join(tier_revenue, by = "Tier") %>%
  mutate(
    Pct_Customers = n_customers / sum(n_customers) * 100,
    Pct_Revenue   = total_revenue / sum(total_revenue) * 100,
    Tier = factor(Tier, levels = c("One-time","Occasional","Regular","Loyal"))
  ) %>%
  pivot_longer(cols = c(Pct_Customers, Pct_Revenue),
               names_to = "Metric", values_to = "Percentage") %>%
  mutate(Metric = recode(Metric,
                         "Pct_Customers" = "% of Customers",
                         "Pct_Revenue"   = "% of Revenue"))

ggplot(tier_summary, aes(x = Tier, y = Percentage, fill = Metric)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.88) +
  scale_fill_manual(values = c("% of Customers" = "#E8002D",
                               "% of Revenue"   = "#222222")) +
  labs(title = "Customer Loyalty Tiers: Share of Customers vs. Revenue",
       x = NULL, y = "Percentage (%)", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```


---

### RQ4 — Inventory Velocity (Fast vs. Slow-moving Products)

For each product, we will compute a **sell-through velocity** = units sold / days active in the dataset. A **box plot** of units sold (y-axis) vs. days active (x-axis), colored by product category, will reveal fast-moving vs. stagnant inventory.

```{r velocity-real, fig.width=7, fig.height=5}
velocity <- transactions_clean %>%
  group_by(article_id) %>%
  summarise(
    units_sold  = n(),
    days_active = as.numeric(max(t_dat) - min(t_dat)) + 1,
    velocity    = units_sold / days_active
  ) %>%
  left_join(articles_clean %>% select(article_id, product_group_name),
            by = "article_id") %>%
  filter(!is.na(product_group_name), days_active > 1)

top_groups <- velocity %>%
  count(product_group_name, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(product_group_name)

velocity %>%
  filter(product_group_name %in% top_groups) %>%
  mutate(product_group_name = recode(product_group_name,
                                     "Underwear" = "Inner Wear")) %>%
  ggplot(aes(x = reorder(product_group_name, velocity, median),
             y = velocity, fill = product_group_name)) +
  geom_boxplot(outlier.shape = NA) +
  coord_flip() +
  scale_y_continuous(limits = quantile(
    velocity$velocity[velocity$product_group_name %in% top_groups],
    c(0.01, 0.95), na.rm = TRUE)) +
  scale_fill_manual(values = c(
  "#FAD4D8",
  "#F5A3AD",
  "#EF6B7A",
  "#E8002D",
  "#A60021"
)) +
  labs(title    = "Sell-Through Velocity by Product Group",
       subtitle = "Units sold per day active — higher = faster moving",
       x = NULL,
       y = "Units Sold per Day",
       fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

## Conclusions

This project examined the behavior of customers and the performance of products in a fashion e-commerce environment using the H&M transaction data. The analysis shows that the purchasing patterns differ across the type of product and group of customers and that some of the departments are often co-occurring in the purchase history of the customer.

We also see that customer loyalty is a major aspect, with a small number of repeat customers providing a considerable portion of the total revenue. Also, different products have varying velocity, which implies that demand varies among the categories.

Such results can be used to shape marketing efforts, targeting customers and inventory choices. The next generation of work might include some extra data like browsing history or stock level to have a more in-depth picture of customer intent and product performance.

---
