---
title: "Customer Behavior and Product Performance Analysis in Fashion E-Commerce"
author: "GOAT & Company"
output: html_document
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(lubridate)
library(forcats)
```

# Introduction

This project analyzes customer behavior and product performance using real-world fashion e-commerce data. The goal is to understand purchasing patterns, customer demographics, repeat purchase behavior, and product performance. These insights can help businesses improve recommendation systems, marketing strategies, and inventory decisions.

# Data Description

The dataset used in this project is from the Kaggle competition:

**H&M Personalized Fashion Recommendations**

Dataset link: https://www.kaggle.com/competitions/h-and-m-personalized-fashion-recommendations/data

This dataset contains real-world fashion retail data from H&M and includes:

- `transactions_train.csv`: customer purchase records (date, customer_id, article_id, price)
- `customers.csv`: demographic information such as age and membership status
- `articles.csv`: product metadata such as product type, category, and department

Due to the large size of the dataset, some analyses may be performed on a sampled subset to improve computational efficiency while preserving overall trends.

# Load Data

```{r load-data}
transactions <- read_csv("transactions_train.csv")
customers <- read_csv("customers.csv")
articles <- read_csv("articles.csv")
```

# Initial Data Cleaning

## Check Missing Values

```{r missing-values}
colSums(is.na(transactions))
colSums(is.na(customers))
colSums(is.na(articles))
```

## Remove Duplicates

```{r remove-duplicates}
transactions <- transactions %>% distinct()
customers <- customers %>% distinct()
articles <- articles %>% distinct()
```

## Convert Datatypes

```{r convert-datatypes}
transactions <- transactions %>%
  mutate(
    t_dat = as.Date(t_dat),
    customer_id = as.character(customer_id),
    article_id = as.character(article_id)
  )

customers <- customers %>%
  mutate(
    customer_id = as.character(customer_id),
    age = as.numeric(age),
    FN = as.factor(FN),
    Active = as.factor(Active),
    club_member_status = as.factor(club_member_status),
    fashion_news_frequency = as.factor(fashion_news_frequency)
  )

articles <- articles %>%
  mutate(
    article_id = as.character(article_id),
    product_code = as.character(product_code),
    product_type_no = as.factor(product_type_no),
    department_no = as.factor(department_no),
    index_code = as.factor(index_code),
    index_group_no = as.factor(index_group_no),
    section_no = as.factor(section_no),
    garment_group_no = as.factor(garment_group_no)
  )
```

## Handle Missing Values

```{r handle-missing}
customers <- customers %>%
  mutate(
    age = ifelse(is.na(age), median(age, na.rm = TRUE), age),
    FN = fct_na_value_to_level(FN, level = "Missing"),
    Active = fct_na_value_to_level(Active, level = "Missing"),
    club_member_status = fct_na_value_to_level(club_member_status, level = "Missing"),
    fashion_news_frequency = fct_na_value_to_level(fashion_news_frequency, level = "Missing")
  )
```

## Standardize Categorical Variables

```{r standardize}
customers <- customers %>%
  mutate(
    club_member_status = str_to_title(str_trim(as.character(club_member_status))),
    fashion_news_frequency = str_to_title(str_trim(as.character(fashion_news_frequency)))
  )

articles <- articles %>%
  mutate(across(where(is.character), ~ str_squish(str_trim(.x))))
```

# Merge Datasets

```{r merge}
hm_data <- transactions %>%
  left_join(customers, by = "customer_id") %>%
  left_join(articles, by = "article_id")
```

# Feature Engineering

## Customer Metrics

```{r features}
hm_data <- hm_data %>%
  group_by(customer_id) %>%
  mutate(
    total_purchases_customer = n(),
    first_purchase_date = min(t_dat),
    last_purchase_date = max(t_dat),
    recency_days = as.numeric(max(t_dat) - t_dat)
  ) %>%
  ungroup()
```

## Age Groups

```{r age-groups}
hm_data <- hm_data %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Under 20",
      age < 30 ~ "20-29",
      age < 40 ~ "30-39",
      age < 50 ~ "40-49",
      age < 60 ~ "50-59",
      age >= 60 ~ "60+",
      TRUE ~ "Unknown"
    )
  )
```

## Loyalty Tiers

```{r loyalty}
hm_data <- hm_data %>%
  mutate(
    loyalty_tier = case_when(
      total_purchases_customer == 1 ~ "One-time",
      total_purchases_customer <= 5 ~ "Occasional",
      total_purchases_customer <= 20 ~ "Regular",
      TRUE ~ "Loyal"
    )
  )
```

# Handle Outliers

```{r outliers}
Q1 <- quantile(hm_data$price, 0.25, na.rm = TRUE)
Q3 <- quantile(hm_data$price, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

hm_data_clean <- hm_data %>%
  filter(price >= Q1 - 1.5 * IQR_value,
         price <= Q3 + 1.5 * IQR_value)
```

# Analysis

## Age Distribution

```{r age-plot}
ggplot(customers, aes(x = age)) +
  geom_histogram(bins = 30) +
  labs(title = "Age Distribution", x = "Age", y = "Count")
```

## Loyalty Tier Distribution

```{r loyalty-plot}
hm_data_clean %>%
  count(loyalty_tier) %>%
  ggplot(aes(x = loyalty_tier, y = n)) +
  geom_col() +
  labs(title = "Customer Loyalty Distribution", x = "Tier", y = "Customers")
```

## Top Products

```{r top-products}
hm_data_clean %>%
  count(prod_name, sort = TRUE) %>%
  head(10) %>%
  ggplot(aes(x = reorder(prod_name, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Products", x = "Product", y = "Sales")
```

## Product Velocity

```{r velocity}
item_velocity <- hm_data_clean %>%
  group_by(article_id) %>%
  summarise(
    total_sales = n(),
    days_active = as.numeric(max(t_dat) - min(t_dat)) + 1,
    velocity = total_sales / days_active
  )

ggplot(item_velocity, aes(x = days_active, y = total_sales)) +
  geom_point(alpha = 0.3) +
  labs(title = "Sales vs Days Active")
```

# Conclusion

This project explored customer purchasing behavior and product performance using H&M transaction data. We identified patterns in customer age distribution, loyalty levels, and product popularity.

The results show that customer segmentation and product performance analysis can provide valuable insights for improving recommendation systems, marketing strategies, and inventory management in fashion e-commerce.